Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-05-27 main.tex
Changes while on the phone with Allen

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
the solution of least-squares problems with varying $A$
matrices\footnote{During the face alignment algorithm presented later, The $A$
matrix changes between alignment iterations but not within the L1 solver}.  In
Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with exact $\ell_1$-min
solutions such as BP and Homotopy methods, and other IST-type
algorithms.  Therefore, we choose ALM as the core algorithm for
implementation of the parallel face recognition pipeline.

@@

Note that in the iterative procedure \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrangian $L_{\mu_k} (\xx,\yy_k)$ compared to solving the original problem
\eqref{eq:l1min} directly. In fact, this problem can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the generic ALM $\ell_1$-min algorithm.\footnote{For conciseness, we
only present the ALM algorithm in the primal domain. There also
exist implementations in the dual domain \cite{YangJ2009,YangA2010-ICIP}.}

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities\footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}.

Our target GPU architecture is the Nvidia Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently\footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}.

@@

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency
\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}.  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

To mitigate the problem, if the system can obtain a good initial guess of the
transformation $\tau$ (\eg, provided by a good face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:alignment_alm}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:alignment_alm}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow (B_j^\dagger)^T \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

Optimal implementation of Algorithm \ref{alg:alignment_alm} on a multi-core CPU must take 
into account the properties of the cache hierarchy. In general, cache that is 
closer to the core (i.e. L1 cache) has higher bandwidth, but smaller size compared to cache
that is farther from the core (L3 cache).  For reference, Figure \ref{fig:cpu_caches} 
shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a representative example of a modern multi-core CPU.

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $B_j$ and
$B_j^\dagger$ during the iterative process to execute Algorithm
\ref{alg:alignment_alm}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

@@

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, such as the shrinkage
operator in step 7 of Algorithm \ref{alg:alignment_alm}, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

@@

Solving one alignment problem per SMP necessitated the implementation of a
custom CUDA routine for computing $B_j^\dagger = (B_j^TB_j)^-1 B^T = G^-1 B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Additionally, reductions on long vectors (i.e. sum or $\ell_1$ norm
of an image) are performed using a double precision temporary variable for accumulation.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

@@

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  All temporary variables are stored in GPU DRAM
during solver execution, with the sole exception of the pseudo-inverse of $B_j$,
which is computed on the CPU via Intel's LAPACK implementation, and then
uploaded to GPU DRAM.

@@

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Runtime vs. Dimensions of $A$.}
\label{fig:random_data}
\end{figure}

\subsection{Face Alignment Stage Benchmark} 
\label{sec:benchmark}
This section presents benchmarks of our CPU and GPU implementations of the
alignment stage of the face recognition pipeline, as shown in Algorithm
\ref{alg:iterative_alignment}, as well as the recognition stage of the
face recognition pipeline, as seen in Algorithm \ref{alg:alm_rec}.

First, in order to test our theory that the GPU performs best solving many
$\ell_1$ problems-per concurrently, we benchmark three implementations of the
alignment $ell_1$-min with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU is averaged over a large number of trials, which are run
sequentially or concurrently depending on the implementation.
The results are shown in Table \ref{tbl:ubench}.
%The results are shown below:
\begin{table}[h] 
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SMP using CUDA streams 
\footnote{The streams implementation is limited significantly by a 
cap on the number of concurrent streams in the current version of CUDA}
& 70\,ms \\
\hline
One problem per SMP using single kernel & {\bf 36\,ms} \\
%\hline
\label{tbl:ubench}
\end{tabular}
\end{table}
Our proposed parallelization of the $\ell_1$-min used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.

For experiments on face data we use subsets of the CMU Multi-PIE Face Database.
For gallery images we use frontal images from session 1, which contains 20
images per user taken under different illuminations. For test images we use
images from session 2.  The training images are prepared as follows:
The iterative alignment stage seeks a similarity
transformation between the coordinate frame of the "camera" resolution test
image and the coordinate frame of the "canonical" resolution images which the
images are compared.  We use a $64 \times 64$ pixel window in the canonical
frame.  The training images are aligned by manually clicking the two outer eye
corners in the camera frame, and then warping the images via a similarity
transformation such that the eye corners lie at a pair of canonical eye
locations in the canonical frame.  The canonical eye corners are chosen such
that for well aligned images, the canonical window contains exclusively pixels
from the subject's face.  We assume the face detector is accurate, so for the test
images we use clicked eye corners for test image alignment initialization\footnote{Previous
publications have shown that iterative alignment improves recognition accuracy substantially
even with clicked eye corners}.

In Figure \ref{fig:alignment_stage_runtime} we report the time it takes fort
our proposed CPU and GPU algorithms to align a single test image against each
of the users in the gallery.  We vary the size of the gallery to show how
implementation scales in the number of gallery users.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_runtime_graph}
\caption{Alignment Stage Runtime vs. Size of Training Database}
\label{fig:alignment_stage_runtime}
\end{figure}
Similar to the results in Figure \ref{fig:random_data},
there is a crossover point above which the GPU out-performs the CPU.  Since
the CPU algorithm solves problems sequentially, there is a linear trend.  
The tiers seen in the GPU profile result from the GPU hardware scheduling alignment
problems in concurrent batches, but the overall trend is linear as expected.

The following experiment fulfills the dual purpose of verifying the accuracy of
the alignment stage, and motivating a threshold on how many users are kept for
the recognition stage.  Setting this threshold establishes a compromise between
recognition speed and potential recognition rate.  To explore the affect of the
threshold on potential recognition rate, we vary the number of users kept for
the recognition stage and estimate the probability of the test user being
among the set of training users kept for the recognition stage.  This curve
bounds above the achievable recognition rate of the overall pipeline.  As can
be seen in Figure \ref{fig:user_alignment_rank_plot}, there are rapidly
diminishing returns as the number of kept users is increased beyond 20; a small
fraction of the gallery.
\begin{figure} 
\centering
\includegraphics[width=3.4in]{figures/user_alignment_rank_plot}
\caption{Probability of the test user being kept for the recognition stage vs.
Number of training users kept for the recognition stage}
\label{fig:user_alignment_rank_plot} \end{figure}

Finally, we perform an experiment verifying the recognition accuracy of the
overall pipeline.  Since would expect higher recognition rates but longer
execution time for higher resolution images, we compare the recognition rate of
the overall pipeline with the face recognition stage run at 5 different
canonical image resolutions: $32\times32, 48 \times 48, 64 \times 64, 96 \times
96, \textrm{and} 128 \times 128$.  As motivated by Figure
\ref{fig:user_alignment_rank_plot}, we fix the number of gallery users to 20,
giving approximately 95\% probability to have the true user in the training
set.  
%The inner loop tolerance to 1e-5 and outer loop tolerance to 1e-2.  
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/accuracyVsResolution}
\caption{Accuracy vs Image Resolution}
\label{fig:accuracy_vs_resolution}
\end{figure}
As can be seen in Figure \ref{fig:accuracy_vs_resolution}, at the optimal
resolution, which happens to match the alignment stage resolution, this bound
is very nearly achieved by the GPU implementation.  For significantly lower
resolutions, the accuracy drops off significantly.  Of the two main stages of
the face recognition pipeline, the iterative alignment stage is by far the most
expensive; the recognition stage requires less than one second.

>>>>

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
the solution of least-squares problems with varying $A$
matrices.\footnote{During the face alignment algorithm presented later, The $A$
matrix changes between alignment iterations but not within the L1 solver}  In
Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while achieving estimation accuracy competitive with exact $\ell_1$-min
solutions such as BP and Homotopy methods, and other IST-type
algorithms.  Therefore, we choose ALM as the core algorithm for
implementation of the parallel face recognition pipeline.

@@

Note that in the iterative procedure \eqref{eq:alm}, the second
step only involves vector algebra and matrix-vector multiplication. Therefore,
the procedure is computationally efficient only if it is easier to minimize the
augmented Lagrangian $L_{\mu_k} (\xx,\yy_k)$ compared to solving the original problem
\eqref{eq:l1min} directly. In fact, this problem can be solved element-wise
iteratively by a soft-thresholding algorithm \cite{WrightS2008,BeckA2009},
whose time complexity is bounded by $O(n^2)$ and can be easily parallelized.
Algorithm \ref{alg:alm} summarizes the generic ALM $\ell_1$-min algorithm. \footnote{For conciseness, we
only present the ALM algorithm in the primal domain. There also
exist implementations in the dual domain \cite{YangJ2009,YangA2010-ICIP}.}

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. \footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}

Our target GPU architecture is the Nvidia Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

While the programming model for
the SIMD units of each SMP is somewhat more flexible than on the CPU,
leveraging the flexibility typically comes at the cost of reduced concurrency
.\footnote{In CUDA terms, code branches that cause warp divergence usually
result in serialization of the different groups of threads}  Therefore, for
the purposes of comparing hardware architectures, the SIMD units on the GPU are
roughly analogous to the CPU SIMD units. Thus, in summary, the GPU hardware 
provides two levels of concurrency: {\em SMP-level}
concurrency and {\em thread-level} concurrency.  Note that the GPU provides more
concurrency than a CPU at both levels (14 SMP's vs. 8 cores) and (64 wide vs. 4 wide SIMD
units).

@@

To mitigate the problem, if the system can obtain a good initial guess of the
transformation $\tau$ (\eg, provided by a good face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

@@

Optimal implementation of Algorithm \ref{alg:iterative_alignment} on a multi-core CPU must take 
into account the properties of the cache hierarchy. In general, cache that is 
closer to the core (i.e. L1 cache) has higher bandwidth, but smaller size compared to cache
that is farther from the core (L3 cache).  For reference, Figure \ref{fig:cpu_caches} 
shows the the sizes of L2 and L3 caches of the Intel E5530 CPU, 
which is a representative example of a modern multi-core CPU.

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  Therefore, if eight instances of
\eqref{eq:alignment-linearization} were be executed in parallel on the eight
cores, there would be an insufficient amount of L1 or L2 cache to hold $B_j$ and
$B_j^\dagger$ during the iterative process to execute Algorithm
\ref{alg:iterative_alignment}. Therefore, the bottleneck in this strategy becomes the
bandwidth between the L2 cache and L3 cache. The implementation of some
previous works \cite{WagnerA2009-CVPR,WagnerA2011-PAMI} is indeed based on this
mapping strategy.

@@

More specifically, in our implementation, most of the operations take
advantage of Intel Math Kernel Library (MKL), a commercial implementation of
the standard Basic Linear Algebra Subprograms (BLAS). For the image resampling
step, we make use of the Intel Integrated Performance Primitives (IPP) library.
Both MKL and IPP are optimized for Intel multi-core CPUs, and are able to
automatically utilize both core-level and vector-level concurrency. For other
operations that are not optimized by Intel in-house libraries, such as the shrinkage
operator in step 7 of Algorithm \ref{alg:iterative_alignment}, we achieve both
vector-level and core-level concurrency through a combination of the Open MP
API \cite{dagum2002openmp} and the automatic vectorization facilities of the
Intel C++ compiler (ICC) \cite{dulong1999overview}. 
%We shall compare the performance of the two parallelism methods in Section \ref{sec:experiment}.

@@

Solving one alignment problem per SMP necessitated the implementation of a
custom CUDA routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Additionally, reductions on long vectors (i.e. sum or $\ell_1$ norm
of an image) are performed using a double precision temporary variable for accumulation.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

@@

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  
All non-scalar temporary variables are stored in GPU DRAM
during solver execution. 
%with the sole exception of the pseudo-inverse of $B_j$,
%which is computed on the CPU via Intel's LAPACK implementation, and then
%uploaded to GPU DRAM.

@@

The results of this benchmark can be found in Figure \ref{fig:random_data}.
The x-axis represents the size of the $A$ matrix and the y-axis represents the
average amount of time it to complete a single problem instance.  The GPU
implementation tends to be faster than the CPU implementation at solving a
single large problem, whereas the CPU implementation is faster at solving a
single small problem.  
\begin{figure}
\begin{center}
\includegraphics[width=3.5in]{figures/time_vs_matrix_size_constant_tol}
\end{center}
\caption{Comparison of $\ell_1$-min runtime vs. dimensions of $A$ on random data.}
\label{fig:random_data}
\end{figure}

\subsection{Face Recognition Pipeline Benchmark} 
\label{sec:benchmark}
This section presents benchmarks of our CPU and GPU implementations of the
alignment stage of the face recognition pipeline, as shown in Algorithm
\ref{alg:iterative_alignment}, as well as the recognition stage of the
face recognition pipeline, as seen in Algorithm \ref{alg:alm_rec}.

First, in order to test our theory that the GPU performs best solving many
$\ell_1$ problems-per concurrently, we benchmark three implementations of the
alignment $\ell_1$-min with $A$ of size $5120 \times 32$, and a fixed
$50$ inner loop iterations for each of $50$ outer loop iterations. The runtime
on a GTX480 GPU is averaged over a large number of trials, which are run
sequentially or concurrently depending on the implementation.
The results are shown in Table \ref{tbl:ubench}.
Our proposed parallelization of the $\ell_1$-min used in the alignment
stage is {\em eight times} faster than an implementation solving a single
problem at a time using the stock BLAS libraries.
%The results are shown below:
\begin{table}[h!]
\caption{Benchmark of implementations of alignment $\ell_1$-min.}
\begin{tabular}{|l|c|}
\hline
Sequential solver using CUBLAS & 302\,ms \\
\hline
One problem per SMP using CUDA streams & 70\,ms  \\
\hline
One problem per SMP using single kernel & {\bf 36\,ms} \\
\hline
\end{tabular}
\label{tbl:ubench}
\end{table}
%\footnote{The streams implementation is limited significantly by a 
%cap on the number of concurrent streams in the current version of CUDA}

For experiments on face data we use subsets of the CMU Multi-PIE Face Database.
For gallery images we use frontal images from session 1, which contains 20
images per user taken under different illuminations. For test images we use
images from session 2.  The training images are prepared as follows:
The iterative alignment stage seeks a similarity
transformation between the coordinate frame of the "camera" resolution test
image and the coordinate frame of the "canonical" resolution images which the
images are compared.  We use a $64 \times 64$ pixel window in the canonical
frame.  The training images are aligned by manually clicking the two outer eye
corners in the camera frame, and then warping the images via a similarity
transformation such that the eye corners lie at a pair of canonical eye
locations in the canonical frame.  The canonical eye corners are chosen such
that for well aligned images, the canonical window contains exclusively pixels
from the subject's face.  We assume the face detector is accurate, so for the test
images we use clicked eye corners for test image alignment initialization.\footnote{Previous
publications have shown that iterative alignment improves recognition accuracy substantially
even with clicked eye corners}

In Figure \ref{fig:alignment_stage_runtime} we report the time it takes fort
our proposed CPU and GPU algorithms to align a single test image against each
of the users in the gallery.  We vary the size of the gallery to show how
implementation scales in the number of gallery users.
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/alignment_runtime_graph}
\caption{Alignment stage runtime vs. size of training database.}
\label{fig:alignment_stage_runtime}
\end{figure}
Similar to the results in Figure \ref{fig:random_data},
there is a crossover point above which the GPU out-performs the CPU.  Since
the CPU algorithm solves problems sequentially, there is a linear trend.  
The tiers seen in the GPU profile result from the GPU hardware scheduling alignment
problems in concurrent batches, but the overall trend is linear as expected.

The following experiment fulfills the dual purpose of verifying the accuracy of
the alignment stage, and motivating a threshold on how many users are kept for
the recognition stage.  Setting this threshold establishes a compromise between
recognition speed and potential recognition rate.  To explore the affect of the
threshold on potential recognition rate, we vary the number of users kept for
the recognition stage and estimate the probability of the test user being
among the set of training users kept for the recognition stage.  This curve
bounds above the achievable recognition rate of the overall pipeline.  As can
be seen in Figure \ref{fig:user_alignment_rank_plot}, there are rapidly
diminishing returns as the number of kept users is increased beyond 20; a small
fraction of the gallery.
\begin{figure} 
\centering
\includegraphics[width=3.4in]{figures/user_alignment_rank_plot}
\caption{Probability of the test user being kept for the recognition stage vs.
number of training users kept for the recognition stage.}
\label{fig:user_alignment_rank_plot} \end{figure}

Finally, we perform an experiment verifying the recognition accuracy of the
overall pipeline.  Since would expect higher recognition rates but longer
execution time for higher resolution images, we compare the recognition rate of
the overall pipeline with the face recognition stage run at 5 different
canonical image resolutions: $32\times32, 48 \times 48, 64 \times 64, 96 \times
96, \textrm{and} 128 \times 128$.  As motivated by Figure
\ref{fig:user_alignment_rank_plot}, we fix the number of gallery users to 20,
giving approximately 95\% probability to have the true user in the training
set.  
%The inner loop tolerance to 1e-5 and outer loop tolerance to 1e-2.  
\begin{figure}
\centering
\includegraphics[width=3.4in]{figures/accuracyVsResolution}
\caption{Recognition rate of the full pipeline vs. face window resolution.}
\label{fig:accuracy_vs_resolution}
\end{figure}
As can be seen in Figure \ref{fig:accuracy_vs_resolution}, at the optimal
resolution, which happens to match the alignment stage resolution, this bound
is very nearly achieved by the GPU implementation.  For significantly lower
resolutions, the accuracy drops off significantly.  Of the two main stages of
the face recognition pipeline, the iterative alignment stage is by far the most
expensive; the recognition stage requires less than one second.
